\section{Reading assignment 3: Astronomical control of solar radiation}

\subsection{Key terms}

\textbf{Plane of the ecliptic}: the plane in which Earth moves around the Sun.

\textbf{Tilt}: The angle at which Earth is tilted away from the line
perpendicular to the plane of its orbit around the sun is $23.5 \degree$. The
tilt angle changes in cycles of 41000 years.

\textbf{Solstices}: The longest and shortest days of the year: June 21, Dec 21.

\textbf{Equinoxes}: The days in March and September when the lengths of night
and day become equal in each hemisphere.

\textbf{Perihelion}: "Close pass", when Earth is 153M km from the Sun. Happens
on January 3rd.

\textbf{Aphelion}: "Distant pass", when Earth is 158M km from the Sun. Happens
on July 4th.

\textbf{Wavelength}: Length of a cycle.

\textbf{Period}: The wavelength of a cycle, expressed in units.

\textbf{Frequency}: The inverse of the period of a cycle.

\textbf{Amplitude}: A measure of the amount by which cycles vary around their
long-term average.

\textbf{Modulation}: Behaviour in which the amplitude of peaks and valleys
changes in a repetitive or cyclic way.

\textbf{Sine waves}: Sinusoidals are perfect cycles because they are regular
both in period and in amplitude (isn't a sawtooth pattern too?).

\textbf{Eccentricity}: The measure of how elliptic (=not perfectly circular)
an ellipse is. It is defined as $\epsilon = \frac{\sqrt{a^2 - b^2}}{a}$.
Earth's orbit's eccentricity is not constant and has varied through time.

\textbf{Axial precession}: Earth's wobbling motion, caused by changing in the
direction of the tilt, at cycle length 25700 years.

\textbf{Precession of the ellipse}: A slow rotation of the entire orbit of the
Earth.

\textbf{Precession of the equinoxes}: The movement of equinoxes (and solstices)
around Earth's orbit which takes around 23000 years to complete. It results
from combined axial precession and precession of the ellipse.

\textbf{Precessional index}: An expression measuring impact of Earth's orbit's
eccentricity and the movement of equinoxes around the orbit. It is defined as
$\epsilon \sin \omega$, where $\epsilon$ is eccentricity and $\omega$ is the
current angle between the Earth-Sun lines at March 20 equinox and at
perihelion. Eccentricity \textit{modulates} the angular motion of the
precession of the equinoxes.

\textbf{Insolation}: Radiation arriving at the top of Earth's atmosphere.

\textbf{Caloric insolation seasons}: The summer caloric half-year is defined as
the 182 days when the incoming insolation exceeds the amount received during
the other 182 days.

\textbf{Time series analysis}: Techniques to extract rhythmic cycles embedded
in the records of climate.

\textbf{Spectral analysis}: One of time series analysis. Gradually sliding a
series of sine waves (with different phases and cycles) and if one has high
correlation with the plot, it is detected as component. Sounds like DFT.

\textbf{Power spectrum}: A type of plot where y-axis represents the amplitude
of the cycles, also known as power.

\textbf{Filtering}: A time series analysis technique, also known as bandpass
filtering.

\textbf{Aliasing}: A term that refers to false trends generated by
undersampling the true complexity in a signal.

\subsection{Review questions}:

\subsubsection{Why does Earth have seasons?}
Because of the tilt in relation to sun. It is summer when a given part of the
world is tilted towards and winter when in opposite direction. Therefore it is
summer in Australia when we have winter in Europe.

\subsubsection{When is Earth closest to the Sun in its present orbit? How does
this "close pass" position affect the amount of radiation received on Earth?}
In its present orbit, Earth is closest to the Sun on January 3. It causes the
winter radiation in the Northern Hemisphere and summer radiation in Southern
Hemisphere to be slightly stronger than they would be in a perfectly circular
orbit.

\subsubsection{Describe in your own words the concept of modulation of a cycle}
Modulation of a cycle are higher level patterns in its amplitude or loudness
if we talk about soundwave or intensity.

\subsubsection{Earth's tilt is slowly decreasing today. As it does so, are the
polar regions receiving more or less solar radiation in summer? In winter?}
As the tilt is decreasing, the Arctic and Antarctic cirles move closer to the
poles. That is, the area with polar day and polar night becomes slower. Polar
regions receive more radiation in the winter and less in the summer.

\subsubsection{How is axial precession different from precession of the
ellipse?}
Axial precession changes the direction of Earth's tilt, precession of the
ellipse moves the Earth's orbit.

\subsubsection{How does eccentricity combine with precession to control a key
aspect of the amount of insolation Earth receives?}
Precession affects how much solar radiation the Earth receives on the
solstices (the closer the solstice to the perihelion, the more radiation) in
a sinewave pattern. Eccentricity modulates this signal.
Since there are multiple cycles of eccentricity changes, with multiple lengths,
they nearly cancel each other out.

\subsubsection{Do insolation changes during summer and winter have the same or
opposite timing at any single location on Earth? Why or why not?}
They have have exactly opposite timing, they are exactly out of phase. This is
because the tilt that brings one pole closer to the Sun, also puts the other
pole farther from the Sun. More intense summer insolation maxima and deeper
winter insolation minima occur together at any one location.

\subsubsection{Do the following changes occur at the same time (same year) in
Earth's orbital cycles?}

\textbf{Summer insolation maxima changes at both poles caused by changes in
tilt?}
No, they are out of phase.

\textbf{Summer insolation maxima in the tropics of both hemispheres by
precession?}
Yes, they are in phase.

\section{Reading assignment 3: Insolation Control of Monsoons}

\textbf{Monsoon creation}: In summer, strong solar radiation causes a rapid and
large warming of the land, but a slower and much less intense warming of the
ocean. Rapid heating over the continents causes air to warm, expand and rise,
and the upward movement of the air creates an area of low pressure at the
surface. The air flowing toward this low-pressure region also warms and rises.
The air arriving from nearby ocean carries water vapor that condenses and
contributes to monsoonal rainfall.

In winter, when solar radiation is weaker, air over the land cools off rapidly,
becomes denser than the air over the still-warm ocean, and sinks from higher
levels in the atmosphere. This downward movement creates a region of high
pressure over the land, in contrast to the lower pressure over the still-warm
oceans. The overall atmospheric flow in winter is a downward-and-outward
movement of cold, dry air from the land to the sea.

\subsection{Key terms}

\textbf{Orbital monsoon hypothesis}: The idea that changes in insolation
control the strength of monsoons over orbital time scales proposed by John
Kutzbach in early 1980s.

\textbf{Nonlinear response}: In this context, the nonlinear response of the
climate system to insolation: the amount of rainfall is highly sensitive to
insolation change in one season (summer) but largely insensitive to changes
in the other (winter).

\textbf{Threshold level}: A threshold level on insolation, below which the
monsoon response will be too weak to leave any evidence in the geologic record.

\textbf{Sanpropels}: Black organic-rich muds, found in Mediterranean sediments.

\textbf{Anoxic}: Lacking the oxygen needed to convert (oxidize) organic carbon
to inorganic form.

\textbf{Clipped response}: The truncation of the summer monsoon response at the
threshold.

\textbf{Orbital tuning}: A method for dating sedimentary records on land and in
the ocean, based on the link between summer insolation forcing and monsoon
responses.

\subsection{Review questions}

\subsubsection{In what way is the orbital monsoon hypothesis an extension of
processes driving modern monsoons?}

Because seasonal monsoon circulations are driven in the modern world by changes
in solar radiation, orbital-scale changes in summer and winter insolation
should have produced a similar response.

\subsubsection{Why does the intensity of 23000-year monsoon peaks vary at
intervals of 100000 and 413000 years?}
Because of the eccentricity changes happening in cycles of 100000 and 413000
years.

\subsubsection{How did the Mediterranean Sea acquire a freshwater lid during
times when very little precipitation was falling in that region?}
The water was delivered by Nile.

\subsubsection{Explain how the opposed July/February timing of past monsoon
changes in China and Brazil lends strong support to the orbital monsoon
hypothesis.}
Cave deposits (stalactites, stalagmites) form as the groundwater drips through
water into the caves. They are constructed of calcite (CaCO$_3$). Their layers
can be accurately dated by radiometric analysis of small amounts of thorium
and uranium. The relative amount of two isotopes of oxygen ($^{16}$O and
$^{18}$O) tells us about precipitation. Layers with highly negative
$\delta^{18}$O indicate strong monsoons.

Caves found in Brazil and China show a very similar pattern of monsoon strength
over the years, and each is in phase with the midsummer in the respective
hemisphere, thus out of phase with each other.

\subsubsection{Does peak monsoon strength lag behind summer insolation
forcing?}
Yes, they lag 2000 years.

\subsubsection{What similarities exist between monsoon changes in Pangea
200 million years ago and those in North Africa during the last several hundred
thousand years?}
Monsoons filled and emptied these Pangaean lakes in response to orbital
precessions in the same way that North African lakes have filled and emptied
during much more recent times.

\subsubsection{How do tectonic uplift and orbital variations combine to affect
the long-term intensity of monsoons?}
Tectonic uplift creates a slow increase in the amplitude of the orbital-scale
cycles. Their combined effect is stronger than a simple linear combination of
the two effects.

\section{Reading assignment 3: Insolation control of ice sheets}

Coral reefs that formed during the last 150000 years confirm that the
$\delta^{18}$O signal is a reasonable proxy for ice sheet size. The ages of the
most prominent $\delta^{18}$O minima correspond to the ages of coral reefs
formed during high stands of sea level caused by ice sheet melting, and the
amplitudes of the sea level changes estimated from the reefs correspond to the
relative changes in ice volume inferred from the $\delta^{18}$O signal.

\subsection{Key terms}

\textbf{Equilibrium line}:
The boundary between positive and negative ice mass balance (that is between
ice sheet growth and ablation), drawn around $10 \degree$.

\textbf{Milankovitch theory}: A systematic way of calculating the impact of
astronomical changes on insolation received on Earth at diferent latitudes and
in different seasons. Formulated by Serbian astronomer Milutin Milankovitch.

Milankovitch proposed that ice growth in the Northern Hemisphere occurs
during times when summer insolation is reduced.

3 factors control the summer insolation:

\begin{itemize}
	\item Earth's tilt
	\item How far from the Sun is the Earth at summer solstice
	\item Orbit's eccentricity
\end{itemize}

Milankovitch noted that the most sensitive latitude for low insolation values
is $65 \degree$N, at which the ice forms first and melts last.

\textbf{Climate point}: The point where the equilibrium line intercepts the
Earth in the higher latitudes. The amount of north-south shift of equilibrium
line is proportional to the amount of change in summer insolation. These shifts
can cover $10 \degree$ to $15 \degree$.

\textbf{Phase lag}: The persistent delay in ice volume relative to summer
insolation. There is also lag in ice volume responses to cycles of orbital
tilt and precession. The lag there is roughly $\frac{1}{4}$ of the cycle
length.

\textbf{Elastic response}: Immediate sinking response of bedrock depression
under the heavy ice load. It represents roughly 30\% of the vertical change
in bedrock.

\textbf{Viscous response}: The slow response of bedrock sinking over thousands
of years. It follows after the elastic response and is caused by the slow flow
of rock in a "soft" layer of the upper mantle at a depth between 100 and 350km.

\textbf{Basal slip}: Meltwater at the base of the ice sheet saturates soft
sediments and creates a lubricated layer across which ice can slide.

\textbf{Termination}: Abrupt deglaciation.

\subsection{Review questions}

\subsubsection{What is the equilibrium line and why is it important?}
The equilibrium line is the boundary between ice sheet growth and ablation.
It is important because after crossing it the ablation accelerates. Ice sheet
doesn't accelerate towards lower temperatures because at very low temperatures
not much snow falls.

\subsubsection{Why are northern ice sheets likely to be more responsive to
insolation changes than ice in Antarctica?}

\subsubsection{Why does the size of a growing or melting ice sheet lag well
behind changes in insolation?}
Both growth and melting is very slow. The rate at which ice grows reaches its
maximum when the summer insolation has reached its minimum. The ice sheet then
grows and continue to grow even as summer insolation is increasing again.
An analogous argument is made about ablation. Even if its rate reaches maximum
at peak summer insolation, it takes time for ablation to accumulate.

\subsubsection{How does the delay in bedrock response to ice loading or
unloading act as a positive feedback on ice volume?}
Ice sheet grows $\rightarrow$ Bedrock sinking delayed for thousands of years
(viscuous response) $\rightarrow$ Ice sheet stays at higher, colder elevations
$\rightarrow$ More positive ice mass balance $\rightarrow$ Ice grows faster

Ice sheet melts $\rightarrow$ Bedrock rebound delayed for thousands of years
(viscuous response) $\rightarrow$ Ice sheet stays at lower, warmer elevations
$\rightarrow$ More negative ice mass balance $\rightarrow$ Ice melts faster

\subsubsection{If the average amplitude of a 41000-year $\delta^{18}$O cycle
in the deep North Atlantic prior to $0.9$ Myr ago was $1.0$‰, and if changes
in ice volume account for half of that total, how large was the average changes
in deep-water temperature?}

\subsubsection{How do corals provide an indication of the volume of water
tied up in ice sheets on land?}
Some coral species grow just below sea levels and have strong structural
frameworks that remain intact long after individual coral organisms have died
and that preserve records of sea level positions. Old coral reefs can be dated
by radiometric decay methods. Their skeletons contain small amounts of
$^{234}$U that slowly decays to $^{230}$Th.

\subsubsection{What has been the average rate of uplift of a coral reef formed
during the last interglaciation 125000 years ago and now lying 131 meters above
sea level?}
$U = \frac{Ht - 6}{125} = \frac{131 - 6}{125} = 1m/1000yr$

\subsubsection{Which part of the Milankovitch theory has proven accurate? Which
parts are insufficient?}
It explains the presence of ice volume responses at 41000 and 23000 years and
why they lag several thousand years behind he summer insolation forcing. But
it fails to explain the dominance of the 41000-year cycle over the 23000-year
signal for almost 2 Myr and it does not account for the emergence of large
oscillations near 100000 years in the last million years.

\subsubsection{If a reef on New Guinea that has been dated to 104000 years
formed at 17m below modern sea levels, and if the average uplift rate on that
island has been 2m per thousand years, what is the current elevation of that
reef?}
$2m = \frac{Ht - 17}{104}$\\
$Ht - 17 = 208m$\\
$Ht = 225m$
